Disentangling linkages between satellite-derived indicators of forest structure and productivity for ecosystem monitoring.

### 2023-11-14 scripts/4_vegan.qmd
resubmit following revisions

# non normalized
loadings_rgb <- clean_loads %>% 
  group_by(axis) %>%
  pivot_wider(names_from = var,
              values_from = axis_loading) %>%
  ungroup() %>%
  replace(is.na(.), 0) %>%
  mutate(across(elev_cv:elev_p95, \(x) ifelse(x > 0.5, 0.5, x))) %>%
  mutate(dhi_rgb = rgb(CumDHI, VarDHI, MinDHI),
         pri_rgb = rgb(elev_p95, percentage_first_returns_above_2m, elev_cv, maxColorValue = 1),
         mod_rgb = rgb(basal_area, total_biomass, gross_stem_volume, maxColorValue = 1)) %>%
  select(strata, axis, percentage_first_returns_above_2m, elev_p95, elev_cv, ends_with("rgb")) %>%
  pivot_longer(ends_with("rgb"))

>>>>

# non normalized
loadings_rgb <- clean_loads_abs %>% 
  group_by(axis) %>%
  pivot_wider(names_from = var,
              values_from = axis_loading) %>%
  ungroup() %>%
  replace(is.na(.), 0) %>%
  mutate(across(elev_cv:elev_p95, \(x) ifelse(x > 0.5, 0.5, x))) %>%
  mutate(dhi_rgb = rgb(CumDHI, VarDHI, MinDHI),
         pri_rgb = rgb(elev_p95, percentage_first_returns_above_2m, elev_cv, maxColorValue = 1),
         mod_rgb = rgb(basal_area, total_biomass, gross_stem_volume, maxColorValue = 1)) %>%
  select(strata, axis, percentage_first_returns_above_2m, elev_p95, elev_cv, ends_with("rgb")) %>%
  pivot_longer(ends_with("rgb"))

### 2023-11-14 scripts/manuscript/manuscript.qmd
resubmit following revisions

Monitoring the changing state of biodiversity globally is key to
understanding and mitigating the increased extinction risk for many
species [@thomas2004; @urban2015] and the homogenization of biotic
communities at various spatial and temporal scales
[@mcgillFifteenFormsBiodiversity2015]. The Group on Earth Observations
Biodiversity Observation Network has developed the Essential
Biodiversity Variables (EBVs) [@pereira2013], which are standardized,
comparable variables that represents a minimum set of biological
information required for monitoring biodiversity change at large spatial
extents [@skidmore2021]. The EBVs are designed as an analog to the
Essential Climate Variables framework [@bojinski2014] to be global in
scope, relevant to biodiversity information, feasible to implement and
use, and complementary to one another [@skidmore2021]. There are six
classes of EBVs collected at various scales. The ecosystem EBV classes
(ecosystem structure and ecosystem function) are well suited to be
monitored across large areas using remote sensing technologies
[@skidmore2021]. This has led to a proliferation of potential EBV
datasets that have been generated to describe these two ecosystem EBV
classes, especially in forested environments.

@@

The suite of remote sensing indicators used in biodiversity monitoring
should be carefully considered. Ideally, indicators should be selected
that are closely related to the phenomenon of interest to allow for
direct integration of monitoring results with management actions
[@pressey2021]. With the advent of large-extent monitoring methods like
satellite remote sensing which have led to a proliferation of potential
EBV datasets, it becomes important to assess the interrelationships
between these datasets and the complementary of the information they
provide to reduce redundancy in future EBV development efforts
[@pereira2013; @skidmore2021]. Identifying and exploring the strong
relationships that may exist between indicators of EBVs may help
elucidate the ecological relationships between these facets of
biodiversity, such as ecosystem structure and function, at local to
global scales. On the other hand, when remote sensing-derived EBVs are
not related, they may be well suited to be used in monitoring programs
together, as complementary sets of observations.

The relationship between ecosystem structure and function have been
examined within a remote sensing context for decades [@huete2002;
@knyazikhin1998; @myneni1994]. Hypothesized mechanisms such as niche
complementarity suggest that productivity increases with stand
structural complexity [@zhang2012], while asymmetric competition for
light can reduce forest productivity with increased structural
complexity [@bourdier2016]. Some aspects of forest structure, such as
canopy cover and leaf area index, also directly influence vegetation
indices, which are used to estimate productivity [@pettorelli2005].
Often, these large-extent estimates of structure and function metrics
are generated from similar data sources, such as the Landsat series of
satellites [@razenkovaMediumresolutionDynamicHabitatInPress ;
@radeloff2019; @matasci2018; @matasci2018b]. Due to this shared data
provenance, and the linkages between biodiversity and both structure and
function, it is important to assess how much unique information these
structure and function metrics represent when used as indicators of
biodiversity.

@@

To obtain representative sample units of the EBV metrics across the
study area, we implemented stratified random sampling across the union
of BEC zones (Table 1) and forest types (i.e., coniferous, broadleaf,
mixed wood, and wetland treed; see @sec-study-area) of BC. Sample units
were restricted to forested pixels that were surrounded by the same
forest type to reduce the chance of sampling mixed pixels. We also
required that sample units were selected from a homogeneous area, to
reduce uncertainty associated with the input datasets [@shang2020];
homogeneous areas were defined as 3x3-pixel windows with a coefficient
of variation for canopy cover and canopy height below 0.5. A minimum
sampling distance of 1-km was implemented to reduce the effects of
spatial autocorrelation. Sample units that had been disturbed in the
last 35 years were discarded by using the disturbance mask generated for
the forested ecosystems of Canada [@hermosilla2016 ] . A maximum of 500
sample units were selected from each stratum; in the many strata where
this was not achieved, all sampling units meeting the above restrictions
were selected. Sampling was conducted in R version 4.2.2 \[@R-base\]
using the **sgsR** package [@R-sgsR]. Neighbourhood analyses for the
land cover classes and coefficient of variations of canopy height and
cover were calculated in Python version 3.9.

@@

Redundancy analysis (RDA) and variation partitioning were used to relate
the primary and modelled forest structure variables to ecosystem
function across a broad environment range. RDA has widely been used in
community ecology where environmental variables of interest are compared
to species composition [@blanchet2014; @kleyer2012]. RDA predicts
multiple response variables by first running a multiple linear
regression of the predictor variables on each response variable, and
then a principal component analysis on the fitted values from each
multiple linear regression. This reduces the dimensionality of the
fitted values by transforming them into a set of independent RDA axes,
and allows the strength of the multivariate relationship between
predictors and responses to be assessed by calculating eigenvalues of
the RDA. Relationships between variables are revealed by the loadings of
both predictor and response variables on the RDA axes. RDA axes are
labelled as RDA1, RDA2, etc. to indicate they are derived from
redundancy analysis, rather than other ordination techniques such as
principal component analysis or constrained correspondence analysis
[@legendre2012]. Partial RDA extends RDA by allowing groups of predictor
variables to be considered as co-variates. Variation partitioning
leverages multiple partial RDAs to assess the overlap between the
explanatory power of the groups of predictor variables [@legendre2012].

We performed RDAs for our full sample, pooling BEC zones and forest
types, as well as individually by BEC zone and forest type, with results
aggregated to BEC zone groups (Table 1) . All measured values were
natural-log transformed and standardized to Z-scores based on the
summary statistics of the full sample prior to analyses. Following the
data transformation, we visually assessed the pairwise relationships
between the overall and stratified datasets for the assumptions of
linearity and homoskedasticity. Forest structure variables were treated
as predictor variables and were grouped based on whether they were
primary estimates from the lidar data or modelled from lidar and
ancillary information (Table 2) for variance partitioning. The three
DHIs were used as the response variables. Analysis of variance tests
were used to determine which RDA axes were significant, with p-values
below 0.1 using an F-test. The proportion of variance attributable to
each axis was calculated from the eigenvalues generated by the RDA, and
axis loadings were calculated as the Pearson correlation between each
forest structure or DHI variable and the RDA axes. To ensure similar
directionality (+ or -) in the loadings across each variable in every
strata, we transformed the data such that the cumulative DHI loading was
always positive. To do so, if the cumulative DHI loading was negative we
multiplied each loading in that strata and axis by -1. Forest structural
attribute axis loadings represent the strength of relationship between a
given variable and the RDA axis, while the DHI loadings indicate what is
being represented by the RDA axes. To visualize the RDA results, we plot
the significant loadings as annotated arrows between the
predictor/response variables and each RDA axis (shown as a box), similar
to a path diagram. We also display the amount of variance explained by
each RDA axis within their respect

@@

By forest type, results indicated similar or increased amounts of
variance explained when compared to the overall dataset
(@fig-fracts-bar). By BEC zones results generally indicated smaller
amounts of variation explained, which may be due to the relatively short
environmental gradients within the zones. The overlap between the
primary and modelled datasets was often responsible for the majority of
the variation explained in the DHIs, except when stratifying by forest
types, where this was only true for the coniferous forests. For
broadleaf, mixed wood, and wetland-treed forest types most of the
explained variation in the DHIs was contributed by the primary forest
structural attributes. A similar pattern was found by BEC zones, where
the variation in the DHIs was principally explained by the overlap
between primary and modelled forest structural attributes, except for
CWH, where no variation was explained by the overlap. The Coastal BEC
group had the lowest amount of variation explained, with Coastal
Douglas-fir having \<1% of variation explained by the structural data.
Overall, the DHIs were decoupled from the forest structural attributes,
as shown by the overall variation explained being under 30% regardless
of data stratification.

```{r}
#| label: fig-fracts-bar 
#| fig-cap: Stacked bar plot proportion of the variation in the Dynamic Habitat Indices (DHIs) variation explained by primary, modelled, and the overlap between primary and modelled structure variables across the stratified and overall datasets.  
knitr::include_graphics(here::here("outputs", "fracts_bar_plot.png"))
```

Stratification by forest type and ecosystem indicated varying loadings
between the RDA axes and predictor/response variables (@fig-radar). All
DHI variables had high loadings on RDA1 (similar to the overall
dataset), with positive cumulative and minimum DHI loadings, and
negative variation DHI loadings. Colder BEC groups (Montane, North, and
Alpine) had smaller minimum DHI loadings, which were half as large as
the other DHI loadings. As such, we identify the first axis as being
consistently associated with overall productivity. Canopy cover
generally had the highest significant predictor loading in RDA1, except
for in non-coniferous forest types (broadleaf, mixed wood and
wetland-treed). These forest types typically had large negative loadings
between RDA1 and structural complexity. The mixed wood forest and
Coastal BEC group had no significant loadings from modelled forest
structural attributes for RDA1. Examining the predictor loadings for
RDA1 spatially across the province shows a geographic pattern in the
loading strengths of primary forest structural attributes (@fig-fcc A).
The interior (Northern and Southern BEC groups) of the province
generally showed high canopy cover loadings, while the Boreal zones in
the northwest had equal loadings in canopy cover and canopy height.
Coastal zones were driven similarly by canopy height and structural
complexity, with very low or non-significant loadings from canopy cover
(@fig-fcc A; @fig-radar). Modelled forest structural attributes for RDA1
generally show similar loadings across the province (@fig-fcc B).

```{r}
#| label: fig-radar
#| fig-cap: Parallel coordinate plots of average absolute significant loading strength by Biogeoclimatic Ecosystem Classification (BEC) group, forest type, and British Columbia. Note the varying y axes.

@@

Only seven of sixteen BEC zones had a second RDA axis, with none being
found in Coastal or Alpine BEC zones (@fig-radar; @fig-fcc). All of the
Montane BEC zones had a secondary axis, while some in the Northern and
Southern interior had a secondary axis. RDA2 often showed the largest
absolute loadings in the minimum DHI, which was frequently, but not
always, negative (@fig-radar; see supplementary material). The montane
group, which had the largest absolute minimum DHI loading, also had the
largest absolute variation DHI loading. The secondary axis had smaller
loadings in the variation and cumulative DHIs, with only the Southern
BEC group having a lower variation DHI loading than cumulative DHI. RDA2
has lower absolute predictor loadings (maximum of \~0.25) when compared
to RDA1 (maximum of \~0.5), with infrequently significant loadings for
canopy cover and structural complexity (@fig-radar). Among the primary
forest structural attributes, RDA2 had the highest loadings with canopy
height in the Southern group, while the Northern group had slightly
larger structural complexity loadings (@fig-radar; @fig-fcc C). Similar
to RDA1, the modelled forest structural attribute loadings were
generally similar to one another (@fig-radar; @fig-fcc D).

```{r}
#| label: fig-fcc
#| fig-cap: False colour maps by Biogeoclimatic Ecosystem Classification (BEC) zone of axis loadings for the first redundancy analysis (RDA) axis (A and B) and second RDA axis (C and D). Colour values are normalized to the maximum loading of each variable. A and C show axis loadings for canopy height (CH; red), canopy cover (CC; green) and structural complexity (SC; blue). B and D show axis loadings for basal area (BA; red), total biomass (TB; green) and gross stem volume (GSV; blue). Black indicates no significant loadings, while greyscale colours indicate similar loadings across the three variables. Annotated colour wheels for each set of variables also shown.

@@

BC shows varying forest structural drivers of the DHIs depending on the
geographic location (@fig-fcc). The interior of BC is drier and warmer,
and also has productivity being strongly driven by canopy cover, while
the Coastal zones are the only BEC zone with strong structural
complexity loadings. The climate of these regions could be influencing
the DHIs [@perez2016], leading to varying loadings strengths depending
on the distance to the coast. For example, water availability has been
shown to be a driver of canopy height [@klein2015]. In these wetter
regions such as the Coastal group, structural variability is more likely
to be present in the vertical, rather than horizontal direction,
especially given the relationship between canopy height and structural
complexity [@atkins2022]. Further, the weakest relationships between
structure and function were found in the Coastal group of BEC zones
(@fig-fracts-bar), which may be due to the consistently high amounts of
canopy cover, and lower amounts of snow found in these zones leading to
potentially saturated NDVI values throughout the year. These consistent
forest structural metrics --- such as the high canopy cover values found
in the Coastal BEC zones --- may also lead to lower amounts of variation
explained by these metrics, leading to them being less informative in
the loadings, and other metrics, such as structural complexity having
higher amounts of variance explained (@fig-radar). These complex
linkages between climate, forest structure, and forest productivity
become further complicated by the extreme environmental and topographic
gradients within BC's ecosystems [@pojar1987; @valentine1978soil], which
may warrant study into the scale of these interactions, with a finer (or
non-climate based) ecoregionalization potentially revealing different
relationships between ecosystem structure and function.

@@

We used redundancy analysis and variation partitioning to assess the
complementarity of two potential EBV datasets - forest structure and the
DHIs. We also separated the forest structure indicator datasets into
primary and modelled structure metrics in order to assess the need to
develop more complex structure variables, or if the primary forest
structural attributes are suitable on their own. We found that the
structure metrics were not strongly related to the DHIs, indicating that
they are suitable to be used together as complementary ecosystem-scale
EBVs when monitoring forest environments. It was also found that
variation in the DHIs explained by the overlap between primary and
modelled structure variables was often higher than the variation
explained by either individually, both in the BC-wide dataset and the
stratified datasets. These ecosystem structure and function indicators
are often attainable at global scales using satellite remote sensing,
and recent advances are allowing them both to be generated at medium
spatial resolutions. We suggest that biodiversity researchers focus on
using forest structural attributes derived directly from the lidar data,
and if needed, use a single modelled forest structural attribute, such
as total aboveground biomass. In addition, we highlight that the usage
of intra-annual summarizations of productivity provide novel information
for biodiversity monitoring when used with forest structural attributes
as complementary EBVs for a holistic monitoring system for forest
ecological integrity.

@@

Figure 3: Stacked bar plot proportion of the variation in the Dynamic
Habitat Indices (DHIs) variation explained by primary, modelled, and the
overlap between primary and modelled structure variables across the
stratified and overall datasets.

Figure 4: Parallel coordinate plots of average absolute significant
loading strength by Biogeoclimatic Ecosystem Classification (BEC) group,
forest type, and British Columbia. Note the varying y axes.

Figure 5: False colour maps by Biogeoclimatic Ecosystem Classification
(BEC) zone of axis loadings for the first redundancy analysis (RDA) axis
(A and B) and second RDA axis (C and D). Colour values are normalized to
the maximum loading of each variable. A and C show axis loadings for
canopy height (CH; red), canopy cover (CC; green) and structural
complexity (SC; blue). B and D show axis loadings for basal area (BA;
red), total biomass (TB; green) and gross stem volume (GSV; blue). Black
indicates no significant loadings, while greyscale colours indicate
similar loadings across the three variables. Annotated colour wheels for
each set of variables also shown.

>>>>

Monitoring the changing state of biodiversity globally is key to
understanding and mitigating the increased extinction risk for many
species [@thomas2004; @urban2015] and the homogenization of biotic
communities at various spatial and temporal scales
[@mcgillFifteenFormsBiodiversity2015]. The Group on Earth Observations
Biodiversity Observation Network has developed the Essential
Biodiversity Variables (EBVs) [@pereira2013], which are standardized,
comparable variables that represent a minimum set of biological
information required for monitoring biodiversity change at large spatial
extents [@skidmore2021]. The EBVs are designed as an analog to the
Essential Climate Variables framework [@bojinski2014] to be global in
scope, relevant to biodiversity information, feasible to implement and
use, and complementary to one another [@skidmore2021]. There are six
classes of EBVs collected at various scales. The ecosystem EBV classes
(ecosystem structure and ecosystem function) are well suited to be
monitored across large areas using remote sensing technologies
[@skidmore2021]. This has led to a proliferation of potential EBV
datasets that have been generated to describe these two ecosystem EBV
classes, especially in forested environments.

@@

The suite of remote sensing indicators used in biodiversity monitoring
should be carefully considered. Ideally, indicators should be selected
that are closely related to the phenomenon of interest to allow for
direct integration of monitoring results with management actions
[@pressey2021]. With the advent of large-extent monitoring methods like
satellite remote sensing which have led to a proliferation of potential
EBV datasets, it becomes important to assess the interrelationships
between these datasets and the complementarity of the information they
provide to reduce redundancy in future EBV development efforts
[@pereira2013; @skidmore2021]. Identifying and exploring the strong
relationships that may exist between indicators of EBVs may help
elucidate the ecological relationships between these facets of
biodiversity, such as ecosystem structure and function, at local to
global scales. On the other hand, when remote sensing-derived EBVs are
not related, they may be well suited to be used in monitoring programs
together, as complementary sets of observations.

The relationships between ecosystem structure and function have been
examined within a remote sensing context for decades [@huete2002;
@knyazikhin1998; @myneni1994]. Hypothesized mechanisms such as niche
complementarity suggest that productivity increases with stand
structural complexity [@zhang2012], while asymmetric competition for
light can reduce forest productivity with increased structural
complexity [@bourdier2016]. Some aspects of forest structure, such as
canopy cover and leaf area index, also directly influence vegetation
indices, which are used to estimate productivity [@pettorelli2005].
Often, these large-extent estimates of structure and function metrics
are generated from similar data sources, such as the Landsat series of
satellites [@razenkovaMediumresolutionDynamicHabitatInPress ;
@radeloff2019; @matasci2018; @matasci2018b]. Due to this shared data
provenance, and the linkages between biodiversity and both structure and
function, it is important to assess how much unique information these
structure and function metrics represent when used as indicators of
biodiversity.

@@

To obtain representative sample units of the EBV metrics across the
study area, we implemented stratified random sampling across the union
of BEC zones (Table 1) and forest types (i.e., coniferous, broadleaf,
mixed wood, and wetland treed; see @sec-study-area) of BC. Sample units
were restricted to forested pixels that were surrounded by the same
forest type to reduce the chance of sampling mixed pixels. We also
required that sample units were selected from a homogeneous area, to
reduce uncertainty associated with the input datasets [@shang2020];
homogeneous areas were defined as 3x3-pixel windows with a coefficient
of variation for canopy cover and canopy height below 0.5. A minimum
sampling distance of 1-km was implemented to reduce the effects of
spatial autocorrelation. Sample units that had been disturbed in the
last 35 years were discarded by using the disturbance mask generated for
the forested ecosystems of Canada [@hermosilla2016]. A maximum of 500
sample units were selected from each stratum; in the many strata where
this was not achieved, all sampling units meeting the above restrictions
were selected. Sampling was conducted in R version 4.2.2 [@R-base] using
the **sgsR** package [@R-sgsR]. Neighbourhood analyses for the land
cover classes and coefficient of variations of canopy height and cover
were calculated in Python version 3.9.

@@

Redundancy analysis (RDA) and variation partitioning were used to relate
the primary and modelled forest structure variables to ecosystem
function across a broad environmental range. RDA has widely been used in
community ecology where environmental variables of interest are compared
to species composition [@blanchet2014; @kleyer2012]. RDA predicts
multiple response variables by first running a multiple linear
regression of the predictor variables on each response variable, and
then a principal component analysis on the fitted values from each
multiple linear regression. This reduces the dimensionality of the
fitted values by transforming them into a set of independent RDA axes,
and allows the strength of the multivariate relationship between
predictors and responses to be assessed by calculating eigenvalues of
the RDA. Relationships between variables are revealed by the loadings of
both predictor and response variables on the RDA axes. RDA axes are
labelled as RDA1, RDA2, etc. to indicate they are derived from
redundancy analysis, rather than other ordination techniques such as
principal component analysis or constrained correspondence analysis
[@legendre2012]. Partial RDA extends RDA by allowing groups of predictor
variables to be considered as co-variates. Variation partitioning
leverages multiple partial RDAs to assess the overlap between the
explanatory power of the groups of predictor variables [@legendre2012].

We performed RDAs for our full sample, pooling BEC zones and forest
types, as well as individually by BEC zone and forest type, with results
aggregated to BEC zone groups (Table 1). All measured values were
natural-log transformed and standardized to Z-scores based on the
summary statistics of the full sample prior to analyses. Following the
data transformation, we visually assessed the pairwise relationships
between the overall and stratified datasets for the assumptions of
linearity and homoskedasticity. Forest structure variables were treated
as predictor variables and were grouped based on whether they were
primary estimates from the lidar data or modelled from lidar and
ancillary information (Table 2) for variance partitioning. The three
DHIs were used as the response variables. Analysis of variance tests
were used to determine which RDA axes were significant, with p-values
below 0.1, using an F-test. The proportion of variance attributable to
each axis was calculated from the eigenvalues generated by the RDA, and
axis loadings were calculated as the Pearson correlation between each
forest structure or DHI variable and the RDA axes. To ensure similar
directionality (+ or -) in the loadings across each variable in every
strata, we transformed the data such that the cumulative DHI loading was
always positive. To do so, if the cumulative DHI loading was negative we
multiplied each loading in that strata and axis by -1. Forest structural
attribute axis loadings represent the strength of relationship between a
given variable and the RDA axis, while the DHI loadings indicate what is
being represented by the RDA axes. To visualize the RDA results, we plot
the significant loadings as annotated arrows between the
predictor/response variables and each RDA axis (shown as a box), similar
to a path diagram. We also display the amount of variance explained by
each RDA axis within their respect

@@

By forest type, results indicated similar or increased amounts of
variance explained when compared to the overall dataset
(@fig-fracts-bar). By BEC zone, results generally indicated smaller
amounts of variation explained, which may be due to the relatively short
environmental gradients within the zones. The overlap between the
primary and modelled datasets was often responsible for the majority of
the variation explained in the DHIs, except when stratifying by forest
types, where this was only true for the coniferous forests. For
broadleaf, mixed wood, and wetland-treed forest types most of the
explained variation in the DHIs was contributed by the primary forest
structural attributes. A similar pattern was found by BEC zones, where
the variation in the DHIs was principally explained by the overlap
between primary and modelled forest structural attributes, except for
CWH, where no variation was explained by the overlap. The Coastal BEC
group had the lowest amount of variation explained, with Coastal
Douglas-fir having \<1% of variation explained by the structural data.
Overall, the DHIs were decoupled from the forest structural attributes,
as shown by the overall variation explained being under 30% regardless
of data stratification.

```{r}
#| label: fig-fracts-bar 
#| fig-cap: Stacked bar plot showing proportion of the variation in the Dynamic Habitat Indices (DHIs) explained by primary, modelled, and the overlap between primary and modelled structure variables across the stratified and overall datasets.  
knitr::include_graphics(here::here("outputs", "fracts_bar_plot.png"))
```

Stratification by forest type and ecosystem indicated varying loadings
between the RDA axes and predictor/response variables (@fig-radar). All
DHI variables had high loadings on RDA1 (similar to the overall
dataset), with positive cumulative and minimum DHI loadings, and
negative variation DHI loadings. Colder BEC groups (Montane, North, and
Alpine) had smaller minimum DHI loadings, which were half as large as
the other DHI loadings. As such, we identify the first axis as being
consistently associated with overall productivity. Canopy cover
generally had the highest significant predictor loading on RDA1, except
for in non-coniferous forest types (broadleaf, mixed wood and
wetland-treed). These forest types typically had large negative loadings
between RDA1 and structural complexity. The mixed wood forest and
Coastal BEC group had no significant loadings from modelled forest
structural attributes for RDA1. Examining the predictor loadings for
RDA1 spatially across the province shows a geographic pattern in the
loading strengths of primary forest structural attributes (@fig-fcc A).
The interior (Northern and Southern BEC groups) of the province
generally showed high canopy cover loadings, while the Boreal zones in
the northwest had equal loadings in canopy cover and canopy height.
Coastal zones were driven similarly by canopy height and structural
complexity, with very low or non-significant loadings from canopy cover
(@fig-fcc A; @fig-radar). Modelled forest structural attributes for RDA1
generally show similar loadings across the province (@fig-fcc B).

```{r}
#| label: fig-radar
#| fig-cap: Parallel coordinate plots of average significant loading strength by Biogeoclimatic Ecosystem Classification (BEC) group, forest type, and British Columbia. Note the varying y axes.

@@

Only seven of sixteen BEC zones had a second RDA axis, with none being
found in Coastal or Alpine BEC zones (@fig-radar; @fig-fcc). All of the
Montane BEC zones had a secondary axis, while some in the Northern and
Southern interior had a secondary axis. RDA2 often showed the largest
absolute loadings in the minimum DHI, which was frequently, but not
always, negative (@fig-radar; see supplementary material). The Montane
group, which had the largest absolute minimum DHI loading, also had the
largest absolute variation DHI loading. The secondary axis had smaller
loadings in the variation and cumulative DHIs, with only the Southern
BEC group having a lower variation DHI loading than cumulative DHI. RDA2
has lower absolute predictor loadings (maximum of \~0.25) when compared
to RDA1 (maximum of \~0.5), with infrequently significant loadings for
canopy cover and structural complexity (@fig-radar). Among the primary
forest structural attributes, RDA2 had the highest loadings with canopy
height in the Southern group, while the Northern group had slightly
larger structural complexity loadings (@fig-radar; @fig-fcc C). Similar
to RDA1, the modelled forest structural attribute loadings were
generally similar to one another (@fig-radar; @fig-fcc D).

```{r}
#| label: fig-fcc
#| fig-cap: False colour maps by Biogeoclimatic Ecosystem Classification (BEC) zone of axis loadings for the first redundancy analysis (RDA) axis (A and B) and second RDA axis (C and D). Colour values are normalized to the maximum loading of each variable. A and C show axis loadings for canopy height (CH; red), canopy cover (CC; green) and structural complexity (SC; blue). B and D show axis loadings for basal area (BA; red), total aboveground biomass (AB; green) and gross stem volume (GSV; blue). Black indicates no significant loadings, while greyscale colours indicate similar loadings across the three variables. Annotated colour wheels for each set of variables also shown.

@@

BC shows varying forest structural drivers of the DHIs depending on the
geographic location (@fig-fcc). The interior of BC is drier and warmer,
and also has productivity being strongly driven by canopy cover, while
the Coastal BEC zone group has very low canopy cover loadings and high
structural complexity loadings (@fig-radar). The climate of these
regions could be influencing the DHIs [@perez2016], leading to varying
loadings strengths depending on the distance to the coast. For example,
water availability has been shown to be a driver of canopy height
[@klein2015]. In these wetter regions such as the Coastal group,
structural variability is more likely to be present in the vertical,
rather than horizontal direction, especially given the relationship
between canopy height and structural complexity [@atkins2022]. Further,
the weakest relationships between structure and function were found in
the Coastal group of BEC zones (@fig-fracts-bar), which may be due to
the consistently high amounts of canopy cover, and lower amounts of snow
found in these zones leading to potentially saturated NDVI values
throughout the year. These consistent forest structural metrics --- such
as the high canopy cover values found in the Coastal BEC zones --- may
also lead to lower amounts of variation explained by these metrics,
leading to them being less informative in the loadings, and other
metrics, such as structural complexity having higher amounts of variance
explained (@fig-radar). These complex linkages between climate, forest
structure, and forest productivity become further complicated by the
extreme environmental and topographic gradients within BC's ecosystems
[@pojar1987; @valentine1978soil], which may warrant study into the scale
of these interactions, with a finer (or non-climate based)
ecoregionalization potentially revealing different relationships between
ecosystem structure and function.

@@

We used redundancy analysis and variation partitioning to assess the
complementarity of two potential EBV datasets - forest structure and the
DHIs. We also separated the forest structure indicator datasets into
primary and modelled structure metrics in order to assess the need to
develop more complex structure variables, or if the primary forest
structural attributes are suitable on their own. We found that the
structure metrics were not strongly related to the DHIs, indicating that
they are suitable to be used together as complementary ecosystem-scale
EBVs when monitoring forest environments. It was also found that
variation in the DHIs explained by the overlap between primary and
modelled structure variables was often higher than the variation
explained by either individually. The exception to this was found in
non-coniferous forest types, Coastal BEC zones, and the BC-wide
analysis. These ecosystem structure and function indicators are often
attainable at global scales using satellite remote sensing, and recent
advances are allowing them both to be generated at medium spatial
resolutions. We suggest that biodiversity researchers focus on using
forest structural attributes derived directly from the lidar data, and
if needed, use a single modelled forest structural attribute, such as
total aboveground biomass. In addition, we highlight that the usage of
intra-annual summarizations of productivity provide novel information
for biodiversity monitoring when used with forest structural attributes
as complementary EBVs for a holistic monitoring system for forest
ecological integrity.

@@

Figure 3: Stacked bar plot showing the proportion of the variation in
the Dynamic Habitat Indices (DHIs) explained by primary, modelled, and
the overlap between primary and modelled structure variables across the
stratified and overall datasets.

Figure 4: Parallel coordinate plots of average significant loading
strength by Biogeoclimatic Ecosystem Classification (BEC) group, forest
type, and British Columbia. Note the varying y axes.

Figure 5: False colour maps by Biogeoclimatic Ecosystem Classification
(BEC) zone of axis loadings for the first redundancy analysis (RDA) axis
(A and B) and second RDA axis (C and D). Colour values are normalized to
the maximum loading of each variable. A and C show axis loadings for
canopy height (CH; red), canopy cover (CC; green) and structural
complexity (SC; blue). B and D show axis loadings for basal area (BA;
red), total aboveground biomass (AB; green) and gross stem volume (GSV;
blue). Black indicates no significant loadings, while greyscale colours
indicate similar loadings across the three variables. Annotated colour
wheels for each set of variables also shown.
